---
title: "Project 2"
author: Furred Flies (Caleb Weis, Eva Wu, Jimin Han)
output: github_document
---

```{r setup, include=FALSE}
library(tidyverse)
knitr::opts_chunk$set(echo = TRUE)
```

This file explains the structure of the repository. This file will include our write-up.

## Repo organization 

* [proposal](proposal) folder. This folder includes the RMarkDown and knitted MarkDown documents which contain the information in this proposal.

* [data_cleaning_description](data_cleaning_description) folder. This folder contains RMarkDown documents used to load, examine, and wrangle data from the RAQSAPI and to load data from other sources, the csv files of data loaded from other sources, as well as the README file consisting of the codebook.

* [air_quality](air_quality) folder. This folder contains an [app.R](air_quality/app.R) document consisting of the code used to build the app and a [data](air_quality/data) folder consisting of the data we used to make visualizations in the app. 

* We have a README.Rmd and README.md file for each folder. These files explain the structure of the repository. The README files in the main directory will include our write-up.

```{r graph}
# experimenting w/ Cali data
ex <- read_csv("air_quality/data/CA1_avg.csv") %>%
  select(-row) %>%
  unique()

input <- c()
input$state <- c("California")
input$pollutant <- "PM2.5"

state_filtered <- ex %>%
  filter(state %in% input$state) %>%
  select(year, quarter, state, year_qtr, input$pollutant) %>%
  unique() %>%
  group_by(year, state) %>%
  summarize(air_qual_year = mean(PM2.5, na.rm = TRUE))

state_filtered %>%
  ggplot(aes(year, air_qual_year, color = state)) +
  geom_line() +
  theme_light() +
  labs(title = paste("Air Quality Across Years in", input$state),
       subtitle = paste("measured by", input$pollutant),
       x = "Year", y = "Air Quality", color = "State")
```
