---
title: "Exploring Air Quality in the US since 1971"
author: Furred Flies (Caleb Weis, Eva Wu, Jimin Han)
output: github_document
---

## High-level goal
 
We have chosen to create an interactive and/or animated spatio-temporal visualization on a topic of interest to us (we have not yet determined whether our visualization will be interactive, animated, or both).

## Loading Dataset

```{r setup, include = FALSE}
knitr::opts_chunk$set(message = FALSE, warning = FALSE, echo = FALSE)
library(RAQSAPI)
library(tidyverse)
library(knitr)
library(stringr)
library(sjmisc)
```


```{r load-data}
kable(head(read_csv("/Users/eva/Desktop/proj2-furred-flies/data/AllStates_overall.csv"), 5))
```

Our data is explained further below and in the README in the data folder. It includes 42 columns (although we will certainly cut this down as soon as we know more about the data), and will include hundreds of thousands of rows once we have loaded all of the necessary data for each state (50 years of air quality data x 50 states + DC is a lot of data).
 
## Description of our goals

Our goal for this project is to explore the change in air quality over time in the United States. We chose this question because air quality is very important for people’s health and reflects a country’s social and industrial trends over time. In order to do this, we will primarily rely upon data from the US EPA Air Quality System API, called RAQSAPI and developed by Clinton Mccrowey, a Physical Scientist with the US EPA. This is a very complex API with concentration data on over 25 different air pollutants, on the county, state, Core Based Statistical Area, air monitoring site, primary quality assurance organization, and monitoring agency levels (the latter three are simply different administrative bodies which collect/oversee air quality data; the third is a geographical unit defined by the US Census Bureau), which includes annual, quarterly, and daily summaries stretching back until the early 1960s. We will use county-level quarterly data from each state on the six components which go into forming an Air Quality Index score (including Ozone, PM2.5, PM10, CO, SO2, and NO2) over the past 50 years, in order to form an accurate picture of how air quality has changed throughout the United States during that time. Each dataset has 42 columns, but we will subset our datasets accordingly to include only the variables we want. The number of rows all vary depending on the time range and the state, but generally it's in the 50,000 - 100,000 range. 

We very seriously considered adding a further dataset portraying the change in health data over time to our visualization, after some extremely valuable feedback from other groups. However, after relentless searching, we were unable to find any sufficiently detailed data that we were able to connect to our air quality analysis. So, we have decided to focus our analysis on an exploration of the change in air quality, and the individual factors that affect air quality, during the years 1971 to 2021. We hope that our final visualization will succeed in illuminating how air quality has changed in the US over time, and can therefore contribute to analyses of changes in other potentially related variables over time, such as asthma prevalence, forest fires, or other climatic/health variables.

Our primary goal is to then use this data in order to construct an interactive and/or animated map of Air Quality in the United States over time. We hope to construct such a map using Shiny, and plan to certainly include a contour map and/or gradient color scale depicting Air Quality scores throughout that time. In addition, we would like to add layers to this map for the individual components which go into forming an Air Quality Index score. Depending on the Shiny app structure, we are considering either a single plot with multiple layers for each factor which goes into forming an air quality score and for the air quality score itself, which can be selected one at a time, or exhibiting multiple panels, each with a plot of a given factor over time, or again, the actual air quality index over time. Users will be able to click a button to select layers to display, or select a panel to observe any given factor or the overarching index. We would like to reiterate that our primary goal is to map air quality in the US over time, and that our main feature would be to include an interactive map that displays the air quality measure over time. We plan to depict the time variable using an interactive slider, which may also include a play button that, when clicked, will animate the graph across the years.


## Weekly Plan of Attack:
 
* Week 1 (May 12-May 18): Load all necessary data (Caleb). Wrangle data into desired structure and calculate AQI Scores from data (Caleb). Create README files & repo structure (Jimin). Edit codebook (Eva). Learn how to use Shiny (everyone).

* Week 2 (May 19-May 25): Construct Shiny shell (Eva). Consider adding layers depicting change in prevalence of potentially contributing factors to poor Air Quality over time (everyone).

* Week 3 (May 26-June 1): Finish up maps (Jimin) and line graphs (Eva). Complete Write-Up and Presentation (using xaringen or Shiny) (everyone). Rehearse Presentation (everyone).

## Repo organization 

* Proposal folder. This folder includes the RMarkDown and knitted MarkDown documents which contain the information in this proposal.

* Data folder. This folder contains RMarkDown documents used to load, examine, and wrangle data from the RAQSAPI and to load data from other sources, and csvs of data loaded from other sources.

* Shiny folder. This folder contains Shiny documents used to construct our data visualization. (We have yet to learn how to use Shiny, so some other documents may be included in this folder as well.)

* We have a README.Rmd and README.md file for each folder. These files explain the structure of the repository. The README files in the main directory will include our write-up.
