---
title: "Air Quality Over Time in the Western US,\n1971-2021"
subtitle: "An Exploration by Caleb Weis, Eva Wu, and Jimin Han"
author: "MACS: Data Visualization"
output: ioslides_presentation
date: '31 May 2022'
runtime: shiny
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = FALSE)
library(shiny)
```

# Background
The Clean Air Act of 1970 was the first to establish National Ambient Air Quality Standards, which have remained essentially unchanged since then.

Despite these standards, and the US' resulting pollution reduction, the exacerbation of climate change and relentless use of fossil fuels in industry and commerce has continued to negatively impact the levels of certain pollutants in the US.

In order to visualize the trend in pollutant levels and overall air quality in the US, we decided to 


## Interactive Plot

```{r load-shiny, echo = FALSE}

shinyAppFile(
  appFile = "shiny/app.R",
  options = list(width = "100%", height = 700)
)

```

## Bullets

- Bullet 1
- Bullet 2
- Bullet 3

## R Output

```{r cars}
summary(cars)
```


